---
title: "1. Population Person-Days"
author: "Melissa Bradley"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: simplex
    toc: yes
    toc_float: 
      collapsed: true
editor_options: 
  chunk_output_type: console
---

```{r setup, message=FALSE, warning=FALSE}

# Packages

library(here)
library(tidyverse)   # Tidying
library(ggplot2)     # Plotting
library(ggbreak)     # Y-Axis Breaks
library(scales)      # Plot Formatting
library(sf)          # Shapefiles (Used for Census data here)
library(lubridate)   # Handling Date/Time
library(kableExtra)  # Kable tables
library(DT)          # HTML table rendering (JS DataTables library)
```

# Purpose

This document details the methods used to calculate and visualize the person-days of exposure to PM2.5 from wildfire smoke (WFS) in each Alaskan census tract, by category.

# Data Sources

The following data sources were obtained and used for this part of the analysis:

* 2020 US Census data
* 2003-2020 Alaskan Wildfire-Specific daily PM2.5 estimates, aggregated by census tract (modeled for this study)
* Wildfire-burned acreage data by year from Alaska Fire Service 

```{r import, results='hide'}

setwd("C:/Users/mkbradley2/OneDrive - University of Alaska/Documents/NIH Wildfires and Birth Outcomes")

here::i_am("OneDrive Files/5. Population Exposure/Scripts/1. Population Person-Days.Rmd")

daily_pm_avg <- read.csv(here::here("OneDrive Files/5. Population Exposure/Data/Daily_Average_WFS_PM25_by_Census_Tract.csv"))

data_dir <- here::here("Documents/Fairbanks Birth Exploration/Data/Tracts2020")
shapefile_path <- file.path(data_dir, "Tracts2020.shp")

if (!file.exists(shapefile_path)) {
    url <- "https://live.laborstats.alaska.gov/cen/maps/gis/Tracts2020.zip"
    destfile <- here::here("OneDrive Files/5. Population Exposure/Data/Tracts2020.zip")
    download.file(url, destfile, mode = "wb")
    unzip(destfile, exdir = data_dir)
}
census_tracts <- st_read(shapefile_path)
acres_burned <- read.csv(here::here("OneDrive Files/5. Population Exposure/Data/acres_burned.csv"))

acres_burned <- acres_burned %>%
  rename(Year = year)

```

# Wildfire Smoke Exposure Levels

We defined wildfire smoke exposure levels based on the EPA’s current Air Quality Index standards for fine particulate matter (PM2.5):

* Good AQI: PM2.5 levels 0.0–9.0 µg/m³, minimal health risk.
* Moderate AQI: PM2.5 levels 9.1–35.4 µg/m³, moderate health risk, especially for sensitive groups.
* Unhealthy and higher AQI: PM2.5 levels >35.5 µg/m³, substantial health risks.

To address potential overestimation of background atmospheric components, such as biogenic emissions 
or sea salt aerosols misclassified as wildfire-specific in the GEOS-Chem modeling process, 
a detection threshold of 2 µg/m³ was set for wildfire-specific PM. As a result, 
the Good AQI level was adjusted to produce the following final smoke exposure categories:

* Light: PM2.5 levels 2.0–9.0 µg/m³
* Moderate: PM2.5 levels 9.1–35.4 µg/m³
* Dense: PM2.5 levels >35.5 µg/m³

# Person-Days Calculation

For each census tract and day, we calculated person-days for each exposure level by multiplying each census tract’s population (TOTALPOP) by each level’s indicator variable. We then added a new variable, “Year”, by extracting the year from each date in the dataset. The data was then pivoted to a long format to simplify grouping. Finally, we summed the person-days in each smoke exposure level by year.

This gives us the following annual person-day metrics:

* No_Smoke_PD: Person-days with no PM2.5 exposure.
* Light_PD: Person-days with Light exposure.
* Moderate_PD: Person-days with Moderate smoke exposure.
* Dense_PD: Person-days with Dense exposure.
* Any_Smoke_PD: Person-days with any PM2.5 exposure.

We then added a new variable, "Year", by extracting the year from each date in the dataset. 
The data was then pivoted to a long format to simplify grouping.
Finally, the data was aggregated by year and smoke category, summing up person-days for each category.

# Data Cleaning and Merging

```{r person days, results='hide'}

# Step 1: Join and preprocess data
pm_census <- daily_pm_avg %>%
  left_join(census_tracts, by = "NAME") %>%
  select(-ends_with(".y")) %>%
  rename_with(~ str_remove(., "\\.x$"), ends_with(".x")) %>%
  mutate(
    No_Smoke = ifelse(PM2.5 <= 0.9, 2, 0),
    Light = ifelse(PM2.5 >= 2 & PM2.5 <= 9, 1, 0), 
    Moderate = ifelse(PM2.5 > 9 & PM2.5 < 35.4, 1, 0),
    Dense = ifelse(PM2.5 >= 35.5, 1, 0),
    Any_Smoke = ifelse(PM2.5 >= 2, 1, 0), # Using 2 as cutpoint for Any Smoke here
    Year = year(Date),
    Month = lubridate::month(Date),  # Extract Month
    Week = lubridate::isoweek(Date)  # Extract Week Number
  )

# Step 2: Add person-days calculations
pm_census <- pm_census %>%
  rowwise() %>%
  mutate(
    No_Smoke_PD = No_Smoke * TOTALPOP,
    Light_PD = Light * TOTALPOP,
    Moderate_PD = Moderate * TOTALPOP,
    Dense_PD = Dense * TOTALPOP,
    Any_Smoke_PD = Any_Smoke * TOTALPOP
  ) %>%
  ungroup()

# Step 3: Reshape and summarize by year
pm_census_summary <- pm_census %>%
  pivot_longer(
    cols = c(No_Smoke_PD, Light_PD, Moderate_PD, Dense_PD, Any_Smoke_PD),
    names_to = "Smoke_Category",
    values_to = "Person_Days"
  ) %>%
  mutate(Smoke_Category = str_remove(Smoke_Category, "_PD")) %>%  # Clean category names
  group_by(Year, Smoke_Category) %>%
  summarise(Total_Person_Days = sum(Person_Days, na.rm = TRUE), .groups = "drop")

# Step 4: Monthly summary setup
pm_census_monthly <- pm_census %>%
  mutate(YearMonth = format(as.Date(Date), "%Y-%m")) %>%
  pivot_longer(
    cols = c(No_Smoke_PD, Light_PD, Moderate_PD, Dense_PD, Any_Smoke_PD),
    names_to = "Smoke_Category",
    values_to = "Person_Days"
  ) %>%
  mutate(Smoke_Category = str_remove(Smoke_Category, "_PD")) %>%  # Clean category names
  group_by(YearMonth, Smoke_Category) %>%
  summarise(Total_Person_Days = sum(Person_Days, na.rm = TRUE), .groups = "drop")


```

# Graphs

## Acres Burned + PM - All Years/Yearly

### Acres Burned + PM Graph Setup

```{r graph setup, results='hide'}

# Helper function to calculate y-axis limits
calculate_y_limits <- function(person_days_data, acres_data) {
  # Calculate primary y-axis range based on Person-Days data
  primary_max <- max(person_days_data$Person_Days, na.rm = TRUE) / 1e6
  primary_buffer <- 0.1 * primary_max
  ylim.prim <- c(0, primary_max + primary_buffer)  # Add 10% buffer
  
  # Calculate secondary y-axis range based on Acres Burned data
  secondary_max <- max(acres_data$acres, na.rm = TRUE) / 1e6
  secondary_buffer <- 0.1 * secondary_max
  ylim.sec <- c(0, secondary_max + secondary_buffer)  # Add 10% buffer
  
  list(ylim.prim = ylim.prim, ylim.sec = ylim.sec)
}

# Helper function for dual-axis transformation
normalize_acres <- function(acres, primary_lim, secondary_lim) {
  TEMP <- acres / 1e6  # Acres Burned in millions
  fit <- lm(b ~ . + 0, tibble::tribble(
    ~a, ~s,  ~b,
    1, (secondary_lim[1] - mean(TEMP)) / sd(TEMP), primary_lim[1],
    1, (secondary_lim[2] - mean(TEMP)) / sd(TEMP), primary_lim[2]
  ))
  list(a = fit$coefficients['a'], s = fit$coefficients['s'], TEMP = TEMP)
}

# Helper function to generate the plot
plot_person_days_vs_acres <- function(data, acres_burned, title, y_breaks = NULL) {
  limits <- calculate_y_limits(data, acres_burned)
  norm_params <- normalize_acres(acres_burned$acres, limits$ylim.prim, limits$ylim.sec)
  
  ggplot() +
    # Bar plot for acres burned
    geom_bar(data = acres_burned, aes(
      x = Year, 
      y = norm_params$a + ((acres / 1e6 - mean(norm_params$TEMP)) / sd(norm_params$TEMP)) * norm_params$s
    ), 
    stat = "identity", fill = "skyblue", alpha = 0.6) +
    
    # Line plot for Person-Days
    geom_line(data = data, aes(
      x = Year, 
      y = Person_Days / 1e6, 
      group = Smoke_Category,
      color = Smoke_Category,
      linetype = Smoke_Category
    ), linewidth = 1.5) +
    
    # Point plot for Person-Days
    geom_point(data = data, aes(
      x = Year, 
      y = Person_Days / 1e6, 
      group = Smoke_Category,
      color = Smoke_Category,
      shape = Smoke_Category
    ), size = 4) +
    
    # Unified legend aesthetics
    scale_color_manual(
      values = c("Any Smoke" = "black", "Moderate" = "darkorange", "Dense" = "red"),
      breaks = c("Light", "Moderate", "Dense"),  # Order the legend manually
      name = "Smoke Density:"
    ) +
    scale_shape_manual(
      values = c("Any Smoke" = 16, "Moderate" = 15, "Dense" = 17),
      breaks = c("Light", "Moderate", "Dense"),  # Order the legend manually
      name = "Smoke Density:"
    ) +
    scale_linetype_manual(
      values = c("Any Smoke" = "solid", "Moderate" = "dotted", "Dense" = "dashed"),
      breaks = c("Light", "Moderate", "Dense"),  # Order the legend manually
      name = "Smoke Density:"
    ) +
    
    # Secondary Y-axis for Acres Burned
    scale_y_continuous(
      name = "Person-Days of Exposure (in millions)",
      labels = scales::comma_format(),
      limits = limits$ylim.prim,
      sec.axis = sec_axis(~ (. - norm_params$a) / norm_params$s * sd(norm_params$TEMP) + mean(norm_params$TEMP),
                          name = "Acres Burned (in millions)", labels = scales::comma_format(scale = 1))
    ) +
    
    # X-axis settings
    scale_x_continuous(breaks = seq(2003, 2020, by = 1)) +
    
    # Theme adjustments
    theme_minimal() +
    theme(
      legend.position = "top",
      legend.title = element_text(size = 12, face = "bold"),
      legend.text = element_text(size = 12),
      plot.title = element_text(hjust = 0.5, face = "bold", margin = margin(b = 10)),
      axis.title.x = element_text(margin = margin(t = 10)),
      axis.title.y = element_text(size = rel(1.2), margin = margin(r = 15)),
      axis.title.y.right = element_text(size = rel(1.2), margin = margin(l = 15)),
      axis.text = element_text(size = rel(0.9)),
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
      panel.grid = element_blank(),
      panel.border = element_blank(),
      axis.line.x.bottom = element_line(color = "gray", linewidth = 0.7),
      axis.line.y.left = element_line(color = "gray", linewidth = 0.7),
      axis.line.y.right = element_line(color = "gray", linewidth = 0.7)
    ) +
    
    # Labels
    labs(
      # title = str_wrap(title, width = 60),
      x = "Year"
    )
}



# Define titles
any_smoke_title <- "Person-Days of Exposure to PM2.5 from Wildfire Smoke and Acres Burned in Alaska, 2003-2020"
med_Dense_title <- "Person-Days of Exposure to PM2.5 from Wildfire Smoke and Acres Burned in Alaska, 2003-2020"


```

### Any Smoke
<br>
```{r any only, echo = F, fig.width=8, fig.height=6, dpi=300}

# Plot Any Smoke + Acres Burned
pm_census_any <- pm_census_summary %>%
  filter(Smoke_Category %in% c("Any_Smoke")) %>%
  rename(Person_Days = Total_Person_Days) %>%
  mutate(Smoke_Category = ifelse(Smoke_Category == "Any_Smoke", "Any Smoke", Smoke_Category))

plot_person_days_vs_acres(pm_census_any, acres_burned, any_smoke_title)

# Use below code for saving
# any_smoke_plot <- plot_person_days_vs_acres
# ggsave(filename="any_smoke_plot_final.png", dpi = 300, width = 10, height = 7)


```

### Moderate and Dense Smoke
<br>
```{r med Dense, echo = F, fig.width=8, fig.height=6, dpi=300}

# Plot Moderate and Dense Smoke + Acres Burned
pm_census_med_heav <- pm_census_summary %>%
  filter(Smoke_Category %in% c("Moderate", "Dense")) %>%
  rename(Person_Days = Total_Person_Days) 

plot_person_days_vs_acres(pm_census_med_heav, acres_burned, med_Dense_title)
# ggsave(filename="plot_med_hvy_final.png", dpi = 300, width = 10, height = 7)

```

### Any, Moderate and Dense Smoke
<br>
```{r all, echo = F, fig.width=8, fig.height=6, dpi=300}

# Plot Any Smoke, Moderate, and Dense with y-axis break
pm_census_any_med_heav <- pm_census_summary %>%
  filter(Smoke_Category %in% c("Any_Smoke", "Moderate", "Dense")) %>%
  rename(Person_Days = Total_Person_Days) %>%
  mutate(Smoke_Category = ifelse(Smoke_Category == "Any_Smoke", "Any Smoke", Smoke_Category))

plot_person_days_vs_acres(pm_census_any_med_heav, acres_burned, med_Dense_title) +
  theme(
    axis.text = element_text(size = 17),       # axis tick labels
    axis.title = element_text(size = 20),      # axis titles
    legend.text = element_text(size = 20),     # legend labels
    legend.title = element_text(size = 20)     # legend title
  )

# ggsave(filename="pm_acres_any_med_heav.png", dpi = 300, width = 10, height = 7)

```


## Weekly (>2 Million Burned Acre Years)

### Weekly Graph Setup

```{r weekly graph setup}

pm_census_daily_long <- pm_census %>%
     select(Date, No_Smoke_PD, Light_PD, Moderate_PD, Dense_PD, Any_Smoke_PD) %>%  # Include Date
     pivot_longer(cols = starts_with("No_Smoke_PD"):starts_with("Any_Smoke_PD"),
                  names_to = "Smoke_Category",
                  values_to = "Person_Days") %>%
     mutate(
         Smoke_Category = str_remove(Smoke_Category, "_PD"),  # Clean category names
         Year = year(Date)  # Add Year for filtering
     )

# Step 1: Summarize Weekly Person-Days and Filter Out "No_Smoke"
pm_census_weekly_long <- pm_census_daily_long %>%
  mutate(Date = as.Date(Date),  # Ensure Date is in Date format
         Week = floor_date(Date, unit = "week")) %>%  # Group data into weeks
  filter(Smoke_Category != "No_Smoke") %>%  # Exclude "No_Smoke"
  group_by(Week, Smoke_Category) %>%
  summarise(Person_Days = sum(Person_Days, na.rm = TRUE), .groups = "drop")

# Step 2: Identify Wildfire Season Based on "Light" Readings
season_bounds <- pm_census_weekly_long %>%
  filter(Smoke_Category == "Light", Person_Days > 0) %>%  # Filter for "Light" readings with Person_Days > 0
  mutate(Year = year(Week)) %>%  # Extract year from Week
  group_by(Year) %>%
  summarise(
    Season_Start = min(Week),  # First week with "Light" readings
    Season_End = max(Week),    # Last week with "Light" readings
    .groups = "drop"
  )

# Step 3: Filter Data Within the Wildfire Season
pm_census_weekly_filtered <- pm_census_weekly_long %>%
  mutate(Year = year(Week)) %>%
  left_join(season_bounds, by = "Year") %>%  # Join season bounds
  filter(Week >= Season_Start & Week <= Season_End)  # Keep weeks within wildfire season

# Step 4: Single Year Plotting

plot_weekly_graph <- function(data, year, title= "Alaska Wildfire Smoke Exposures by PM Density,", filter_categories = NULL) {
    # Clean data by removing non-finite values
    data_filtered <- data %>%
        filter(year(Week) == year,  # Filter by year
               Smoke_Category %in% c("Light", "Moderate", "Dense"),
               if (!is.null(filter_categories)) Smoke_Category %in% filter_categories else TRUE,  # Filter smoke categories if specified
               !is.na(Person_Days), is.finite(Person_Days))  # Ensure valid data
    
    # Dynamically set x-axis limits
    x_min <- min(data_filtered$Week, na.rm = TRUE)
    x_max <- max(data_filtered$Week, na.rm = TRUE)
    
    # Create the plot
    ggplot(data_filtered, 
           aes(x = Week, y = Person_Days / 1e3, fill = Smoke_Category)) +
        geom_area(position = "stack", alpha = 0.6, color = "white") +
        scale_fill_manual(
            values = c("Light" = "yellow", "Moderate" = "orange", "Dense" = "red"),
            breaks = c("Light", "Moderate", "Dense"),  # Order the legend manually
            name = "Smoke Density:"  # Add legend title
        ) +
        theme_minimal() +
        theme(
            legend.position = "top",
            legend.title = element_text(size = 12, face = "bold"),
            legend.text = element_text(size = 12),
            plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
            axis.title = element_text(size = 14),
            axis.text = element_text(size = 12),
            panel.grid.major = element_blank(),
            panel.grid.minor = element_blank(),
            plot.margin = unit(c(1, 1, 1, 1), "cm"),
            axis.line.x.bottom = element_line(color = "gray", linewidth = 0.7),
            axis.line.y.left = element_line(color = "gray", linewidth = 0.7),
            axis.line.y.right = element_line(color = "gray", linewidth = 0.7)
        ) +
        labs(
            title = paste(title, year),
            x = "Year",
            y = "Person-Days of Exposure (in thousands)"
        ) +
        scale_y_continuous(labels = scales::comma_format(), limits = c(0, 6000)) +  # Adjust Y scale
        scale_x_date(date_labels = "%b", date_breaks = "1 month")  # Fixed X scale
}

  
# Step 5: All Years Plotting Function
  
plot_weekly_all_years <- function(data, title = "Alaska Wildfire Smoke Exposures by PM Density, 2003-2020", filter_categories = NULL) {
    # Filter and clean data to only include wildfire season and valid categories
    data_filtered <- data %>%
        filter(
            Smoke_Category %in% c("Light", "Moderate", "Dense"),  # Focus on smoke categories
            if (!is.null(filter_categories)) Smoke_Category %in% filter_categories else TRUE,
            !is.na(Person_Days), is.finite(Person_Days)        # Ensure valid data
        ) %>%
      mutate(Year = as.factor(year(Week)))  # Convert year to a factor for y-axis
    
    # Create the plot
    ggplot(data_filtered, 
           aes(x = Week, y = Person_Days / 1e3, fill = Smoke_Category)) +
        geom_area(position = "stack", alpha = 0.6, color = "white") +
        scale_fill_manual(
          values = c("Light" = "yellow", "Moderate" = "orange", "Dense" = "red"),
          breaks = c("Light", "Moderate", "Dense"),  # Order the legend manually
          name = "Smoke Density:"  # Add legend title
        ) +
        theme_minimal() +
        theme(
            legend.position = "top",
            legend.title = element_text(size = 12, face = "bold"),
            legend.text = element_text(size = 12),
            plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
            axis.title = element_text(size = 14),
            axis.text = element_text(size = 12),
            panel.grid.major = element_blank(),
            panel.grid.minor = element_blank(),
            plot.margin = unit(c(1, 1, 1, 1), "cm"),
            axis.line.x.bottom = element_line(color = "gray", linewidth = 0.7),
            axis.line.y.left = element_line(color = "gray", linewidth = 0.7),
            axis.line.y.right = element_line(color = "gray", linewidth = 0.7)
        ) +
        labs(
            title = paste(title),
            x = "Year",
            y = "Person-Days of Exposure (in thousands)"
        ) +
        scale_x_date(date_breaks = "1 year", date_labels = "%Y")  # Breaks by year
}


```

### 2003 (Moderate + Dense)

```{r weekly 2003 graph med Dense, echo = F, fig.width=8, fig.height=6, dpi=300}

plot_weekly_graph(pm_census_weekly_filtered, 2003, filter_categories = c("Moderate", "Dense"))

# For saving
# ggsave(filename="plot_weekly_2003_medDense.png", dpi = 300, width = 10, height = 7)

```

### 2004 (Moderate + Dense)

```{r weekly 2004 graph med Dense, echo = F, fig.width=8, fig.height=6, dpi=300}

plot_weekly_graph(pm_census_weekly_filtered, 2004, filter_categories = c("Moderate", "Dense"))

# For saving
# ggsave(filename="plot_weekly_2004_medDense.png", dpi = 300, width = 10, height = 7)

```

### 2005 (Moderate + Dense)

```{r weekly 2005 graph med Dense, echo = F, fig.width=8, fig.height=6, dpi=300}

plot_weekly_graph(pm_census_weekly_filtered, 2005, filter_categories = c("Moderate", "Dense"))

# For saving
# ggsave(filename="plot_weekly_2005_medDense.png", dpi = 300, width = 10, height = 7)

```

### 2006 (Moderate + Dense)

```{r weekly 2006 graph med Dense, echo = F, fig.width=8, fig.height=6, dpi=300}

plot_weekly_graph(pm_census_weekly_filtered, 2006, filter_categories = c("Moderate", "Dense"))

# For saving
# ggsave(filename="plot_weekly_2006_medDense.png", dpi = 300, width = 10, height = 7)

```

### 2007 (Moderate + Dense)

```{r weekly 2007 graph med Dense, echo = F, fig.width=8, fig.height=6, dpi=300}

plot_weekly_graph(pm_census_weekly_filtered, 2007, filter_categories = c("Moderate", "Dense"))

# For saving
# ggsave(filename="plot_weekly_2007_medDense.png", dpi = 300, width = 10, height = 7)

```

### 2008 (Moderate + Dense)

```{r weekly 2008 graph med Dense, echo = F, fig.width=8, fig.height=6, dpi=300}

plot_weekly_graph(pm_census_weekly_filtered, 2008, filter_categories = c("Moderate", "Dense"))

# For saving
# ggsave(filename="plot_weekly_2008_medDense.png", dpi = 300, width = 10, height = 7)

```

### 2009 (Moderate + Dense)

```{r weekly 2009 graph med Dense, echo = F, fig.width=8, fig.height=6, dpi=300}

plot_weekly_graph(pm_census_weekly_filtered, 2009, filter_categories = c("Moderate", "Dense"))

# For saving
# ggsave(filename="plot_weekly_2009_medDense.png", dpi = 300, width = 10, height = 7)

```

### 2010 (Moderate + Dense)

```{r weekly 2010 graph med Dense, echo = F, fig.width=8, fig.height=6, dpi=300}

plot_weekly_graph(pm_census_weekly_filtered, 2010, filter_categories = c("Moderate", "Dense"))

# For saving
# ggsave(filename="plot_weekly_2010_medDense.png", dpi = 300, width = 10, height = 7)

```

### 2011 (Moderate + Dense)

```{r weekly 2011 graph med Dense, echo = F, fig.width=8, fig.height=6, dpi=300}

plot_weekly_graph(pm_census_weekly_filtered, 2011, filter_categories = c("Moderate", "Dense"))

# For saving
# ggsave(filename="plot_weekly_2010_medDense.png", dpi = 300, width = 10, height = 7)

```

### 2012 (Moderate + Dense)

```{r weekly 2012 graph med Dense, echo = F, fig.width=8, fig.height=6, dpi=300}

plot_weekly_graph(pm_census_weekly_filtered, 2012, filter_categories = c("Moderate", "Dense"))

# For saving
# ggsave(filename="plot_weekly_2012_medDense.png", dpi = 300, width = 10, height = 7)

```

### 2013 (Moderate + Dense)

```{r weekly 2013 graph med Dense, echo = F, fig.width=8, fig.height=6, dpi=300}

plot_weekly_graph(pm_census_weekly_filtered, 2013, filter_categories = c("Moderate", "Dense"))

# For saving
# ggsave(filename="plot_weekly_2013_medDense.png", dpi = 300, width = 10, height = 7)

```

### 2014 (Moderate + Dense)

```{r weekly 2014 graph med Dense, echo = F, fig.width=8, fig.height=6, dpi=300}

plot_weekly_graph(pm_census_weekly_filtered, 2013, filter_categories = c("Moderate", "Dense"))

# For saving
# ggsave(filename="plot_weekly_2013_medDense.png", dpi = 300, width = 10, height = 7)

```

### 2015 (Moderate + Dense)

```{r weekly 2015 graph med Dense, echo = F, fig.width=8, fig.height=6, dpi=300}

plot_weekly_graph(pm_census_weekly_filtered, 2015, filter_categories = c("Moderate", "Dense"))

# For saving
# ggsave(filename="plot_weekly_2015_medDense.png", dpi = 300, width = 10, height = 7)

```

### 2016 (Moderate + Dense)

```{r weekly 2016 graph med Dense, echo = F, fig.width=8, fig.height=6, dpi=300}

plot_weekly_graph(pm_census_weekly_filtered, 2016, filter_categories = c("Moderate", "Dense"))

# For saving
# ggsave(filename="plot_weekly_2016_medDense.png", dpi = 300, width = 10, height = 7)

```

### 2017 (Moderate + Dense)

```{r weekly 2017 graph med Dense, echo = F, fig.width=8, fig.height=6, dpi=300}

plot_weekly_graph(pm_census_weekly_filtered, 2017, filter_categories = c("Moderate", "Dense"))

# For saving
# ggsave(filename="plot_weekly_2017_medDense.png", dpi = 300, width = 10, height = 7)

```

### 2018 (Moderate + Dense)

```{r weekly 2018 graph med Dense, echo = F, fig.width=8, fig.height=6, dpi=300}

plot_weekly_graph(pm_census_weekly_filtered, 2018, filter_categories = c("Moderate", "Dense"))

# For saving
# ggsave(filename="plot_weekly_2018_medDense.png", dpi = 300, width = 10, height = 7)

```

### 2019 (Moderate + Dense)

```{r weekly 2019 med Dense, echo = F, fig.width=8, fig.height=6, dpi=300}

plot_weekly_graph(pm_census_weekly_filtered, 2019, filter_categories = c("Moderate", "Dense"))

# For saving
# ggsave(filename="plot_weekly_2019_medDense.png", dpi = 300, width = 10, height = 7)

```


### 2020 (Moderate + Dense)

```{r weekly 2020 graph med Dense, echo = F, fig.width=8, fig.height=6, dpi=300}

plot_weekly_graph(pm_census_weekly_filtered, 2020, filter_categories = c("Moderate", "Dense"))

# For saving
# ggsave(filename="plot_weekly_2020_medDense.png", dpi = 300, width = 10, height = 7)

```

### 2003-2020 (Moderate + Dense)

```{r weekly 2003 2020 graph med Dense, echo = F, fig.width=8, fig.height=6, dpi=300}

plot_weekly_all_years(pm_census_weekly_filtered, filter_categories = c("Moderate", "Dense"))

# For saving
# ggsave(filename="plot_weekly_all_medDense.png", dpi = 300, width = 10, height = 7)

```

### 2003-2020 Grid Plot (Moderate and Dense)

```{r weekly 2003 2020 grid, echo = F, fig.width=8, fig.height=6, dpi=300}

# Define function for creating the faceted plot with dynamic x-axis limits
plot_facet_grid <- function(data) {
  
# Clean data to avoid missing/non-finite values
  data_filtered <- data %>%
    filter(Smoke_Category %in% c("Moderate", "Dense"),
           !is.na(Person_Days), is.finite(Person_Days))  # Ensure valid data
  # Dynamically calculate season bounds for each year
  season_bounds <- data_filtered %>%
    group_by(Year) %>%
    summarise(
      x_min = min(Week, na.rm = TRUE),  # Start of the season for this year
      x_max = max(Week, na.rm = TRUE),  # End of the season for this year
      .groups = "drop"
    )
  
  # Use ggplot2 + facet_wrap with dynamic scales="free_x"
  ggplot(data_filtered, 
         aes(x = Week, y = Person_Days / 1e3, fill = Smoke_Category)) +
    geom_area(position = "stack", alpha = 0.6, color = "white") +
    scale_fill_manual(
      values = c("Moderate" = "orange", "Dense" = "red"),
      breaks = c("Moderate", "Dense"),
      name = "Smoke Density:"
    ) +
    facet_wrap(~Year, ncol = 3, scales = "free_x") +  # Allow dynamic x-axis scaling
    theme_minimal() +
    theme(
      strip.text = element_text(size = 12, face = "bold"),  # Year labels
      legend.position = "top",
      legend.title = element_text(size = 14),
      legend.text = element_text(size = 12),
      axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
      axis.text.y = element_text(size = 12),
      axis.title = element_text(size = 14),
      plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
      axis.title.y = element_text(size = 18, margin = margin(r = 20)),  
      panel.spacing = unit(1, "lines")
    ) +
    labs(
      # title = "Alaska Wildfire Smoke Exposures by PM Density - High Acres Burned Years (2003-2020)",
      x = "Month",
      y = "Person-Days of Exposure (in Thousands)"
    ) +
    scale_y_continuous(labels = scales::comma_format()) +
    scale_x_date(date_labels = "%b", date_breaks = "1 month")  # Show monthly labels
}

# Apply function to filtered data and save the plot
facet_20032020 <- plot_facet_grid(pm_census_weekly_filtered)
facet_20032020

# Save the final plot
# ggsave("Grid_PM25_2003_2020.png", plot = facet_20032020, dpi = 300, width = 16, height = 10)


```


### High WF Year Grid Plot (Moderate and Dense)

```{r high wf year grid, echo = F, fig.width=8, fig.height=6, dpi=300}

pm_census_weekly_high_wf <- pm_census_weekly_filtered %>%
  filter(Year %in% c(2004, 2005, 2009, 2015, 2019))

facet_highwf <- plot_facet_grid(pm_census_weekly_high_wf)  + 
  theme(
    axis.text.x = element_text(size = 20),
    axis.text.y = element_text(size = 20),
    axis.title.x = element_blank(),  # ✅ Remove "Month"
    axis.title.y = element_text(size = 20, margin = margin(r = 20)),
    legend.text = element_text(size = 20),
    legend.title = element_text(size = 20),
    strip.text = element_text(size = 20),
    panel.spacing = unit(1.5, "lines") 
  )

facet_highwf

# Save the final plot
# ggsave("Grid_PM25_High_WF.jpeg", plot = facet_highwf, dpi = 300, width = 16, height = 10)

```


# First 9 Years and Last 9 Years Comparison

Here we compared wildfire-specific PM2.5 levels observed between 2003-2011 and compared it with 
PM2.5 levels observed between 2012-2020. 

## T-Test

We performed t-tests comparing statewide PM2.5 averages, as well as Light, Moderate, 
and Dense person-days of wildfire smoke exposure for 2003–2011 and 2012–2020. 
There was a statistically significant decrease in statewide wildfire-specific PM2.5 averages from 2003-2011 to 2012-2020 (t(823116)=15.03, 𝑝<0.001). 
There was not a statistically significant difference in Moderate and Dense person-days of wildfire smoke exposure from 2003-2011 to 2012-2020. 

```{r t tests}

# T-Tests----

## Raw PM (Not Person-Days)----

  # Define the years for each period
  first_9_years <- 2003:2011
  last_9_years <- 2012:2020

  pm_census_first_9_years <- pm_census %>% filter(Year %in% first_9_years)
  pm_census_last_9_years <- pm_census %>% filter(Year %in% last_9_years)

  t_test_result_raw_pm <- t.test(pm_census_first_9_years$`PM2.5`, pm_census_last_9_years$`PM2.5`)

  t_test_summary_pm <- data.frame(
    Statistic = t_test_result_raw_pm$statistic,
    "Mean (First 9 Years)" = t_test_result_raw_pm$estimate[1],
    "Mean (Last 9 Years)" = t_test_result_raw_pm$estimate[2],
    "p-value" = t_test_result_raw_pm$p.value,
    "Confidence Interval Lower" = t_test_result_raw_pm$conf.int[1],
    "Confidence Interval Upper" = t_test_result_raw_pm$conf.int[2]
  )

  kable(t_test_summary_pm, caption = "T-test Results for Total Wildfire PM (First 9 Years vs Last 9 Years)")

## Person-Days of Wildfire Smoke Exposure----

exposure_levels <- pm_census_summary %>%
  filter(Smoke_Category %in% c("Light", "Moderate", "Dense")) %>%
  mutate(Period = ifelse(Year %in% first_9_years, "First 9 Years", "Last 9 Years"))

t_test_summary_pd <- exposure_levels %>%
  group_by(Smoke_Category) %>%
  summarise(
    t_test = list(
      t.test(
        Total_Person_Days[Period == "First 9 Years"],
        Total_Person_Days[Period == "Last 9 Years"]
      )
    ),
    .groups = "drop"
  ) %>%
  rowwise() %>%
  mutate(
    Statistic = t_test$statistic,
    `Mean (First 9 Years)` = t_test$estimate[1],
    `Mean (Last 9 Years)` = t_test$estimate[2],
    `p-value` = t_test$p.value,
    `Confidence Interval Lower` = t_test$conf.int[1],
    `Confidence Interval Upper` = t_test$conf.int[2]
  ) %>%
  select(Smoke_Category, Statistic, `Mean (First 9 Years)`, `Mean (Last 9 Years)`, `p-value`, `Confidence Interval Lower`, `Confidence Interval Upper`)

# Display the results

kable(
  t_test_summary_pd,
  caption = "T-test Results for Light, Moderate, and Dense Person-Days of Smoke Exposure (First 9 Years vs Last 9 Years)"
)

```

## Census Tracts Increase/Decrease Table

Below is a table summarizing the total, Moderate, and Dense person-days of wildfire smoke exposure by census tract. Following this, a second table summarizes the number of census tracts that experienced increases or decreases in total person-days of exposure, as well as exposure to Moderate and high smoke densities.

```{r census tract table, echo = F, warning=F}

# Compute detailed changes for each Census Tract
census_change <- pm_census %>%
  group_by(NAME) %>%
  summarise(
    First_9_Years_Total = sum(No_Smoke_PD[Year %in% 2003:2011], na.rm = TRUE),
    Last_9_Years_Total = sum(No_Smoke_PD[Year %in% 2012:2020], na.rm = TRUE),
    Total_Change = Last_9_Years_Total - First_9_Years_Total,
    First_9_Years_Moderate = sum(Moderate_PD[Year %in% 2003:2011], na.rm = TRUE),
    Last_9_Years_Moderate = sum(Moderate_PD[Year %in% 2012:2020], na.rm = TRUE),
    Moderate_Change = Last_9_Years_Moderate - First_9_Years_Moderate,
    First_9_Years_Dense = sum(Dense_PD[Year %in% 2003:2011], na.rm = TRUE),
    Last_9_Years_Dense = sum(Dense_PD[Year %in% 2012:2020], na.rm = TRUE),
    Dense_Change = Last_9_Years_Dense - First_9_Years_Dense,
    Status = ifelse(Total_Change > 0, "Increased", "Decreased")
  ) %>%
  arrange(desc(Total_Change))

# Create the detailed data table
census_table <- datatable(
  census_change,
  options = list(
    pageLength = 10,
    autoWidth = TRUE,
    scrollX = TRUE # Adds horizontal scrolling
  ),
  caption = "Detailed Census Tracts with Person-Days Change Over Periods"
)

# Summary table broken down by Total, Moderate, and Dense changes
summary_change <- census_change %>%
  summarise(
    Total_Days_Increased = sum(ifelse(Total_Change > 0, 1, 0)),
    Total_Days_Decreased = sum(ifelse(Total_Change <= 0, 1, 0)),
    Moderate_Days_Increased = sum(ifelse(Moderate_Change > 0, 1, 0)),
    Moderate_Days_Decreased = sum(ifelse(Moderate_Change <= 0, 1, 0)),
    Dense_Days_Increased = sum(ifelse(Dense_Change > 0, 1, 0)),
    Dense_Days_Decreased = sum(ifelse(Dense_Change <= 0, 1, 0))
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "Category",
    values_to = "Count"
  )

# Create the summary data table
summary_table <- datatable(
  summary_change,
  options = list(
    pageLength = 5,
    autoWidth = TRUE
  ),
  caption = "Summary of Census Tracts by Status of Change (Total, Moderate, Dense)"
)

# Display the tables
census_table
summary_table

```

# Start and End Dates for Unhealthy Smoke By Year

This code identifies the start and end dates for unhealthy levels of smoke exposure in Alaska for each year in the dataset.

```{r start end dates}

# Step 2: Filter for high wildfire years
high_wildfire_years <- c(2004, 2005, 2009, 2013, 2015, 2019)

pm_census_filtered <- pm_census %>%
  filter(Year %in% high_wildfire_years)

# Step 3: Summarize person-days for Moderate and Dense exposure levels
exposure_summary <- pm_census_filtered %>%
  select(Year, Month, Week, Moderate, Dense, Any_Smoke) %>%
  group_by(Year, Month, Week) %>%
  summarize(
    Moderate_Days = sum(Moderate, na.rm = TRUE),
    Dense_Days = sum(Dense, na.rm = TRUE),
    Any_Days = sum(Any_Smoke, na.rm = TRUE),
    .groups = "drop"
  )

start_end_summary <- exposure_summary %>%
  pivot_longer(cols = c(Any_Days, Moderate_Days, Dense_Days), names_to = "Exposure_Level", values_to = "Days") %>%
  filter(Days > 0) %>%
  group_by(Year, Exposure_Level) %>%
  summarize(
    start_week = min(Week),
    end_week = max(Week),
    .groups = "drop"
  )

start_end_summary <- start_end_summary %>%
  mutate(
    start_date = as.Date(paste(Year, start_week, 1), format = "%Y %U %u"),
    end_date = as.Date(paste(Year, end_week, 1), format = "%Y %U %u"),
    start_date = format(start_date, "%m/%d/%Y"),  # Convert to MM/DD/YYYY format
    end_date = format(end_date, "%m/%d/%Y"),
    start_month = month(as.Date(paste(Year, start_week, 1), format = "%Y %U %u"), label = TRUE, abbr = FALSE),  # Full month name
    end_month = month(as.Date(paste(Year, end_week, 1), format = "%Y %U %u"), label = TRUE, abbr = FALSE)   # Full month name
  )
print(start_end_summary)

```
